Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-04 paper/main.tex
Updates

In the DAG literature, the primary focus has been on developing automated
algorithms to learn causal structures from datasets. These efforts have led to
numerous causal discovery algorithms, including constraint-based methods like
the PC algorithm \citep{Spirtes2001} and Fast Causal Inference (FCI)
\citep{Spirtes2000}), score-based methods such as Hill-Climb Search and Greedy
Equivalence Search (GES) \citep{Chickering2002}, and continuous
optimization-based methods like NO TEARS \citep{Zheng2018} and DAGMA
\citep{Bello2022}. While DAG-based methods focus on automated discovery,
SEM-based methods emphasize expert-driven model specification. This includes
tools to assist researchers in manually constructing or modifying models,
enabling experts to incorporate their domain knowledge in the model building
process. Researchers typically begin with an initial model based on their
expertise and then use these tools to guide modifications that improve the
model's fit to data. This process is commonly known as \emph{Specification
Search} \citep{Long1983} and uses method such as modification indices, and
Wald-based tests \citep{Marcoulides2018}.

@@

\begin{enumerate}
	\item \textbf{Lack of Trust:} While most algorithms are asymptotically
		consistent, their behavior on finite samples is not well
		understood. Additionally, their output can vary significantly
		depending on the choice of algorithm and hyperparameters,
		making it difficult to assess reliability. The absence of
		robust performance evaluation methods further reduce the
		confidence in their outputs. As a result, even when using these
		algorithms are used, researchers often rely on post hoc manual
		modifications to the model based on model testing and their
		domain expertise.
	\item \textbf{Outputs Markov Equivalence Class (MEC):} As multiple
		DAGs can be faithful to a given observational dataset, the
		algorithms can only recover the MECs. These MECs can contain a
		combination of directed and undirected edges. However, most
		methods for downstream tasks, such as identification or causal
		effect estimation, assume knowledge of a fully oriented DAG. As
		a result, researchers have to manually orient these edges based
		on their expert knowledge before using them for analysis.
\end{enumerate}

@@

\caption{A comparison of Markov Equivalence Classes (MEC) learned using Adult Income Dataset \citep{Becker1996} using (a \& b) PC algorithm using a mutual information based test, (c \& d) Hill Climb Search with Bayesian Information Criterion (BIC) score, (e \& f) PC algorithm with a residualization based test \citep{Ankan2023}. The learned model structure varies significantly depending on the algorithm, the CI test, and the sample size used.}
\label{fig:intro}
\end{figure}

Our main contributions in this paper are as follows:
\begin{enumerate}
	\item We present a measure of (conditional) association for mixed data
		based on canonical correlations. This measure of association
		generalizes the widely used partial correlation to mixed data
		setting.
	\item Using this partial association measure, we present a procedure to
		rank CI tests. This ranking can be used to prioritize addition
		or deletion of edges from the model.
	\item We present both a web-browser based tool and an implementation in
		the Python package to allow researchers to apply this method to
		their own datasets.
\end{enumerate}

The rest of the paper is structured as follows. In
Section~\ref{sec:background}, we give a background on the measures of
association for various data types. In Section~\ref{sec:mixed_association}, we
present our method for computing mixed data association. In
Section~\ref{sec:modification} and Section~\ref{sec:web}, we present how to use
this method to make modifications and the web tool, and lastly in
Section~\ref{sec:empirical}, we show some empirical analysis on how well this
modification process works.

>>>>

In the DAG literature, the primary focus has been on developing automated
algorithms to learn causal structures from datasets. These efforts have led to
numerous causal discovery algorithms, including constraint-based methods like
the PC algorithm \citep{Spirtes2001} and Fast Causal Inference (FCI)
\citep{Spirtes2000}), score-based methods such as Hill-Climb Search and Greedy
Equivalence Search (GES) \citep{Chickering2002}, and continuous
optimization-based methods like NO TEARS \citep{Zheng2018} and DAGMA
\citep{Bello2022}. While DAG-based methods focus on automated discovery,
SEM-based methods emphasize expert-driven model specification. This includes
tools to assist researchers in manually constructing models, enabling them to
incorporate their domain knowledge in the model building process. Researchers
typically begin with an initial model based on their expertise and then use
these tools to guide modifications that improve the model's fit to data. This
process is commonly known as \emph{Specification Search} \citep{Long1983} and
uses method such as modification indices, and Wald-based tests
\citep{Marcoulides2018}.

@@

\begin{enumerate}
	\item \textbf{Lack of Trust:} While most algorithms are asymptotically
		consistent, their behavior on finite samples is not well
		understood. Additionally, their output can vary significantly
		depending on the choice of algorithm and hyperparameters,
		making it difficult to assess reliability. The absence of
		robust performance evaluation methods for any given dataset
		further reduce the confidence in their outputs. 
		% As a result,
		% even when using these algorithms are used, researchers often
		% rely on post hoc manual modifications to the model based on
		% model testing and their domain expertise.
	\item \textbf{Outputs Markov Equivalence Class (MEC):} As multiple
		DAGs can be faithful to a given observational dataset, automated 
		algorithms can only recover the MECs. These MECs can contain a
		combination of directed and undirected edges. However, most
		methods for downstream tasks, such as identification or causal
		effect estimation, assume knowledge of a fully oriented DAG. 
		% As
		% a result, researchers have to manually orient these edges based
		% on their expert knowledge before using them for analysis.
\end{enumerate}

@@

\caption{A comparison of Markov Equivalence Class (MEC) learned using Adult Income Dataset \citep{Becker1996} using (a \& b) PC algorithm using a mutual information based test, (c \& d) Hill Climb Search with Bayesian Information Criterion (BIC) score, (e \& f) PC algorithm with a residualization based test \citep{Ankan2023}. The learned model structure varies significantly depending on the algorithm, the CI test, and the sample size used.}
\label{fig:intro}
\end{figure}

Our main contributions in this paper are as follows:
\begin{enumerate}
	\item We present a measure of conditional association for mixed data
		based on canonical correlations. This measure of association
		generalizes some of the commonly used special case measures of 
		association to mixed data.
	\item Using this conditional association measure, we present a procedure to
		manually construct DAGs by ranking violation of implied CIs in the 
		model. This iterative procedure allows users to integrate their
		domain knowledge in the model construction procedure.
	\item Lastly, to allow researchers to easily apply this method on their
		datasets, we present a web-browser based interactive tool as well
		as implementation in a Python package.
\end{enumerate}

The rest of the paper is structured as follows. In
Section~\ref{sec:background}, we give a background on the commonly used
measures of association for various data types. In
Section~\ref{sec:mixed_association}, we present our generalized measure of
conditional association for mixed data. Section~\ref{sec:modification} provides
details on the procedure for using the measure of association for constructing
DAGs. Section~\ref{sec:web} presents our web-browser based tool and lastly in
Section~\ref{sec:empirical}, we show some empirical results to compare this
manual approach to automated algorithms.
